animbook: Visualizing changes in performance measures and demographic affiliations using animation

### 2024-02-14 animbook-journal.Rmd
08/02/2024 to do list

  The structure of this paper is as follows. The next section explains the animation in The New York Times article and why it is relevant to the problem of studying zombie companies. It is followed by sections that describe the data format needed, considerations about making animations, and visual design. The software and its functions and usage are detailed in the next section. The next section illustrates the usage of the software to examine company performance, particularly zombie patterns, and changing political allegiance of voters in Australia.
The discussion section summarises, outlines some limitations and suggests some future directions.

@@

```{r nyt, fig.cap="Screenshot of the New York Times animation, which is the motivation for this visualization package.", fig.width=8, fig.align='center', out.width="100%", layout = "l-body"}
knitr::include_graphics("figures/NYT.png")
```

@@

  The New York Times visualization is also similar to what is called a Sankey plot. Sankey plots are flow diagrams that show the movement/change from one category levels to another, with the width of each arrow representing the proportional changes between levels. There are several R packages available to generate version of Sankey plots including \CRANpkg{ggalluvial} and \CRANpkg{ggparallel}. The main difference between the Sankey plot visualization and the animation featured in the New York Times lies in their methods of representing the proportion of changes. Sankey plots utilize the width of connecting lines, while the New York Times animation employs the density and color of points.

@@

```{r animation-diagram, fig.cap="An animation requires small changes shown in quick succession to make it appear as if the objects are in motion. The three plots above show a single point changing position by a small amount, which when shown quickly will look like that point has moved up and to the right.", fig.width=8, fig.align='center', out.width="100%", layout = "l-body"}
knitr::include_graphics("figures/animation-diagram.png")
```

@@

```{r animated-diagram, fig.cap="Frames 1 and 2 show the year 1 and year 2 positions of points, from the data in the table at top. The id is necessary to enable interpolation to make the small changes to generate a perception of motion from frame 1 position to frame 2 position.", fig.width=8, fig.align='center', out.width="100%", layout = "l-body"}
knitr::include_graphics("figures/animated-diagram.png")
```

@@

```{r proportional-shade, fig.show="hold", out.width="50%", fig.cap="The plot shows the difference between the sigmoid (left) and the sine curve (right). In the sigmoid, as the curve progresses it appears to become narrower, possibly resulting in more misperception of relative proportions compared to the sine curve."}
knitr::include_graphics("figures/sigmoid-shade.png")
knitr::include_graphics("figures/sine-shade.png")
```

  In the animation, the use of proportional shaded areas serves to enhances the comprehension of proportion information, effectively displaying the distribution within each group. To represent sub-groups, points are colored to focus attention on differences between groups of observation.

  Discerning the movement pattern or proportion of sub-groups can be challenging when there is only a small number of points. To address this, sample size are inflated by calculating the sub-group proportion and multiplying it by the desired total number of points. This approach ensures a richer representation of movement patterns.

 Another issue is that all observations have a common starting time. This would mean all points start the animation simultaneously and move as a block through time. This poses a challenge in distinguishing and tracking individual observations and their change patterns. To alleviate this, the animation randomly starts observations at different times so a continuous flow of points is shown. It is important to note that this does not necessarily reflect the real-time occurrence of observations.

  The paths of observations are also interpolated to create small changes in the position of points to produce the appearance of motion. The interpolation functions in \CRANpkg{gganimate} and \CRANpkg{plotly} are inadequate because they only generate a linear path. Our animation needs non-linear path to show the flow. Sigmoid curves are commonly used in Sankey diagrams (@Hvitfeldt_2018). However, @Shaffer_2019 argues that a sine curve more accurately represents proportion throughout the path. As illustrated in Figure \@ref(fig:proportional-shade). The sigmoid accurately represents the proportion at the beginning and end, but as it curves, the shape gets narrower, leading to a less accurate proportion representation than the sine curve. New interpolation functions are needed to compute both the sigmoid and sine path.

@@

  The `time_dependent` argument provides the user flexibility to choose whether points will start simultaneously (`time_dependent = TRUE`) or randomly (`time_dependent = FALSE`). The challenge with points starting at the same time is the formation of point clusters, making it difficult to track individual points. Allowing for the random starting time makes the point tracking simpler. The starting times for the points are generated using the `runif()` function, where the `max` argument can be adjusted using the `width` parameter in the plotting function.

@@

  The names of these plots draws inspiration from a native Australian animals. The comprehensive plot is named Kangaroo, while its subset is referred to as Wallaby, representing a smaller version of the Kangaroo. The final plot is names after Funnel Web Spider, drawing parallels between the plot's features and the distinctive legs of the spider.

@@

```{r shade-algorithm, fig.show="hold", out.width="50%", fig.cap="The plot shows how the algorithm for the proportional\\_shade() function works. The left figure represents the initial step of the algorithm, which calculates all the corner points. The right figure demonstrates the subsequent step, where points in-between the left and right are interpolated using the sine() function."}
knitr::include_graphics("figures/sankey-shade-1.png")
knitr::include_graphics("figures/sankey-shade-2.png")
```

  The algorithm behind the `proportional_shade()` function for generating the proportional shaded data is illustrated in Figure \@ref(fig:shade-algorithm). It started by calculating the corner points for all of the shaded areas. Then, use the `sine()` function to interpolate the point between the left and right.

@@

The `cat_change` data in the package is used to illustrate how to use this package. It has five categories, A to E, over two time points. Since the data is already in the categorized data structure, the `anim_prep_cat()` function is used. The output from this function is a class called `categorized`.

@@

Figure \@ref(fig:top-many), show the flexibility to use various options, such as `subset` or `relation`, to customize the plot appearance. Here `subset = "top"` is used to specify category A for the final time point, and `relation = "many_one"` is used to specify that all categories flowing into A at time point 2 are shown.

@@

Using `subset = "B"` will only show the point for a single category as shown in the Figure \@ref(fig:b-many).

@@

  @oecd_report reported that companies in the United States have a faster metabolic rate relative to Japanese companies. This means that US companies tends to move up and down performance metrics more frequently than Japanese companies. This can be examine using the wallaby and kangaroo plots. The wallaby plot allows the study of entry and exit behavior. The kangaroo plot allows studying the full movement of companies over time.

@@

```{r kan-osiris-figure, fig.cap = "Four frames from the kangaroo plot visualization showing the movement of the Japanese and US companies between the performance sales quantiles from 2006 to 2018 for a sample of data extracted from the Osiris database. The 'not listed' category indicates companies not yet listed or removed from the listing. Most companies stay in the same quantile group, with a small number moving up and down. Most of the movement is made by American companies, which supports the OECD report that US companies have higher turnover rate relative to Japanese companies.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/osiris.png")
```

```{r kan-osiris-gif, out.extra="class = 'gif'", fig.cap = "The kangaroo plot visualization shows the movement of the Japanese and US companies between the performance sales quantiles from 2006 to 2018 for a sample of data extracted from the Osiris database. The 'not listed' category indicates companies not yet listed or removed from the listing. Most companies stay in the same quantile group, with a small number moving up and down. Most of the movement is made by American companies, which supports the OECD report that US companies have higher turnover rate relative to Japanese companies.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/kan-osiris.gif")
```

In the kangaroo plot (Figure `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{fig:kan-osiris-figure}','\\@ref(fig:kan-osiris-gif)'))`) the five grey bars indicate the five categories (Top 25%, 25-50, 50-75, 75-100, Not listed) across years 2006 and 2018. Points represent individual companies with color indicating country of registration. Points that stay within the bar indicate companies that remain in the same quartile over the entire time period. If a company drop in performance or improve performance to the extent of switching quartile, the corresponding point will jump up or down in this plot. Note that all the companies in the "Not listed" category were listed at some point during this time period. It can be seen that most companies stay in the same quantile group over the time period. There are several companies that move up and down, and most of these are US companies. This support the @oecd_report claim that US companies have a higher turnover rate compared to Japanese companies.

Perceiving the differences between the movement pattern of groups can be difficult, especially when the groups have different sizes, as is the case here. There are more US companies than Japanese companies and particularly there are more US companies in the top 25 percent quantile. Thus there will likely be more blue points (US) moving than red points (JP) purely for this reason. The difference that is of interest is **relative proportion**, that one country has a higher fraction of companies moving between quartiles than the other. This animated Sankey plot requires the reader to mentally calculate the fraction of points that are moving, separately for the blue and red points, to answer the question whether one group has a higher movement rate. The task is easier if the two groups have a similar number of points.

@@

  In Figure `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{fig:osiris-figure}','\\@ref(fig:osiris-gif)'))`, the focus is placed on companies that exited the market in 2018 that were in the market at some point since 2006. The grey regions show the proportion of companies in each of the categories in 2006 on the left that were exited the market by 2018. (explain the different in proportion and not listed companies). In 2006, the companies represented by points, colored by country are in one of the quartiles or "Not listed" category. All the companies in the Top 25% were US companies. Any Japanese companies in the Top 25% in 2006 did not exited the market (Not listed). But actually, there were non Japanese companies in the Top 25% in 2006. It is also interesting that the larger proportion of the companies in the Top 25% are getting de-listed, and the lower the quartiles, the less likely that the companies will exit the market.

  Using the wallaby plot, it can be employed to emphasized the high turnvoer rate observed in the kangaroo plot, as it specifically focuses on a subset, allowing the users to better perceive the **relative proportion**. However, it is important to note that this plot is confined to only two-time points (start and end), which may not provide a comprehensive view of the entire story.

@@

@@

```{r voter-figure, fig.cap = "Four frames from a wallaby plot visualization shows how the top party performs in keeping the old voters of different genders. Most voters remain loyal to the party, but a small fraction of voters with roughly equal male-to-female ratio switch primarily to the other major party. Interestingly, the few individuals who identified as neither male or female overwhelmingly shifted their party affiliations to the Greens.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/animation-voter.png")
```

```{r voter-gif, out.extra="class = 'gif'", fig.cap = "The wallaby plot visualization shows how the top party performs in keeping the old voters of different genders. Most voters remain loyal to the party, but a small fraction of voters with roughly equal male-to-female ratio switch primarily to the other major party. Interestingly, the few individuals who identified as neither male or female overwhelmingly shifted their party affiliations to the Greens.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/voter.gif")
```

@@

  According to the survey (@lgbtiq), the Greens Party provided a detailed response to all nine priority areas that focus on changing systems and addressing health and well-being disparities among LGBTIQ+ communities. However, the Labor Party only give a broad statement of commitment to a range of LGBTIQ+ and human right issues. This includes working with LGBTIQ+ Australians and advocates to ensure equality before the law and full access to Medicare.

@@

Beginning with inspiration drawn from the New York Times article, this package provides tools to generate animations for temporal flow data to study changes in groups over time. The paper describes how to pre-process the data into a format for making several types of plots and animation. It also explains what needs to be considered when making an animation such as persistence of objects, how to achieve this using interpolation, and the visualization design choices automated and those available to users.

  The package provides the tools to create animations for simple temporal patterns, with a few steps, but these visualization methods have some limitations. It is primarily for communication rather than exploration because one typically knows what the main patterns are. If there are many categories (quantiles) the display will be too crowded. Because the animation involves perceiving patterns in groups of moving points it can be challenging with unbalanced sample sizes. This requires the reader to mentally calculate the fraction of points that are in motion.

  There are some potential new directions to develop this work. Some applications might benefit by re-parameterization of the x-axis from years to ages (e.g. company ages, time since the company was founded). This can be done by pre-processing before passing to animbook, but it would be more convenient to automate for the user. One could also imagine re-arranging the mapping of variables to the plot and animations, particularly, split on time, and animate on the quantiles to focus on movement through the ranks. The New York Times article had examples of this idea. The static funnel web plot is a good place to experiment with this, and to further enhance visually.

>>>>

  The structure of this paper is as follows. The next section explains the animation in The New York Times article and why it is relevant to the problem of studying zombie companies. It is followed by sections describing the data format needed, considerations about making animations, and visual design. The software and its functions and usage are detailed in the next section. The next section illustrates the usage of the software to examine company performance, particularly zombie patterns, and the changing political allegiance of voters in Australia. The discussion section summarizes, outlines some limitations, and suggests some future directions.

@@

```{r nyt, fig.cap="Screenshot of the New York Times animation, which is the motivation for this visualization package. The animation illustrates the disparity between white men and black men. It reveals that despite black boys growing up in a rich household, they still have a higher likelihood of falling into lower income groups compared to white boys.", fig.width=8, fig.align='center', out.width="100%", layout = "l-body"}
knitr::include_graphics("figures/NYT.png")
```

@@

  The New York Times visualization is also similar to what is called a Sankey plot. Sankey plots are flow diagrams that show the movement/change from one category level to another, with the width of each arrow representing the proportional changes between levels. There are several R packages available to generate versions of Sankey plots, including \CRANpkg{ggalluvial} and \CRANpkg{ggparallel}. The main difference between the Sankey plot visualization and the animation featured in the New York Times lies in their methods of representing the proportion of changes. Sankey plots utilize the width of connecting lines, while the New York Times animation employs the density and color of points.

@@

```{r animation-diagram, fig.cap="An animation requires small changes shown in quick succession to make it appear as if the objects are in motion. The three plots above show a single point changing position by a small amount, which when shown quickly, will look like that point has moved up and to the right.", fig.width=8, fig.align='center', out.width="100%", layout = "l-body"}
knitr::include_graphics("figures/animation-diagram.png")
```

@@

```{r animated-diagram, fig.cap="Frames 1 and 2 show the year 1 and year 2 positions of points from the data in the table at the top. The id is necessary to enable interpolation to make the small changes to generate a perception of motion from frame 1 position to frame 2 position.", fig.width=8, fig.align='center', out.width="100%", layout = "l-body"}
knitr::include_graphics("figures/animated-diagram.png")
```

@@

```{r proportional-shade, fig.show="hold", out.width="50%", fig.cap="The plot shows the difference between the sigmoid (left) and the sine curve (right). In the sigmoid, as the curve progresses, it appears to become narrower, possibly resulting in more misperception of relative proportions compared to the sine curve."}
knitr::include_graphics("figures/sigmoid-shade.png")
knitr::include_graphics("figures/sine-shade.png")
```

  In the animation, the use of proportional shaded areas serves to enhance the comprehension of proportion information, effectively displaying the distribution within each group. To represent sub-groups, points are colored to focus attention on differences between groups of observation.

  Discerning the movement pattern or proportion of sub-groups can be challenging when there are only a small number of points. To address this, the sample size is inflated by calculating the sub-group proportion and multiplying it by the desired total number of points. This approach ensures a richer representation of movement patterns.

  Another issue is that all observations would have a common starting time. It would mean all points start the animation simultaneously and move as a block through time. It poses a challenge in distinguishing and tracking individual observations and their change patterns. To alleviate this, the observations in the animation are randomly started at different times so a continuous flow of points can be seen. It is important to note that this does not necessarily reflect the real-time occurrence of observations.

  The paths of observations are also interpolated to create small changes in the position of points to produce the appearance of motion. The interpolation functions in \CRANpkg{gganimate} and \CRANpkg{plotly} are inadequate because they only generate a linear path. Our animation needs a non-linear path to show the flow. Sigmoid curves are commonly used in Sankey diagrams (@Hvitfeldt_2018). However, @Shaffer_2019 argues that a sine curve more accurately represents proportion throughout the path. As illustrated in Figure \@ref(fig:proportional-shade). The sigmoid accurately represents the proportion at the beginning and end, but as it curves, the shape gets narrower, leading to a less accurate proportion representation than the sine curve. New interpolation functions are needed to compute both the sigmoid and sine paths.

@@

  The `time_dependent` argument provides the user flexibility to choose whether points will start simultaneously (`time_dependent = TRUE`) or randomly (`time_dependent = FALSE`). The challenge with points starting at the same time is the formation of point clusters, making it difficult to track individual points. Allowing for the random starting time makes point tracking simpler. The starting times for the points are generated using the `runif()` function, where the `max` argument can be adjusted using the `width` parameter in the plotting function.

@@

  The names of these plots draw inspiration from native Australian animals. The comprehensive plot is named Kangaroo, while its subset is referred to as Wallaby, representing a smaller version of the Kangaroo. The final plot is named after the Funnel Web Spider, drawing parallels between the plot's features and the distinctive legs of the spider.

@@

  The `proportional_shade()` function's algorithm begins by calculating the corner points for each shaded area. Subsequently, the algorithm utilizes the `sine()` function to interpolate points between the left and right boundaries. Once this interpolation is completed, all the generated points are interconnected to form the shaded regions.

@@

The `cat_change` data in the package is used to illustrate how to use this package. It has five categories, A to E, over two-time points. Since the data is already in the categorized data structure, the `anim_prep_cat()` function is used. The output from this function is a class called `categorized`.

@@

Figure \@ref(fig:top-many), shows the flexibility to use various options, such as `subset` or `relation`, to customize the plot appearance. Here `subset = "top"` is used to specify category A for the final time point, and `relation = "many_one"` is used to specify that all categories flowing into A at time point 2 are shown.

@@

Using `subset = "B"` will only show the point for a single category as shown in Figure \@ref(fig:b-many).

@@

  @oecd_report reported that companies in the United States have a faster metabolic rate relative to Japanese companies. This means that US companies tend to move up and down performance metrics more frequently than Japanese companies. This can be examined using the wallaby and kangaroo plots. The wallaby plot allows the study of entry and exit behavior. The kangaroo plot allows for studying the full movement of companies over time.

@@

```{r kan-osiris-figure, fig.cap = "Four frames from the kangaroo plot visualization showing the movement of the Japanese and US companies between the performance sales quantiles from 2006 to 2018 for a sample of data extracted from the Osiris database. The 'not listed' category indicates companies not yet listed or removed from the listing. Most companies stay in the same quantile group, with a small number moving up and down. Most of the movement is made by American companies, which supports the OECD report that US companies have a higher turnover rate relative to Japanese companies.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/osiris.png")
```

```{r kan-osiris-gif, out.extra="class = 'gif'", fig.cap = "The kangaroo plot visualization shows the movement of the Japanese and US companies between the performance sales quantiles from 2006 to 2018 for a sample of data extracted from the Osiris database. The 'not listed' category indicates companies not yet listed or removed from the listing. Most companies stay in the same quantile group, with a small number moving up and down. Most of the movement is made by American companies, which supports the OECD report that US companies have a higher turnover rate relative to Japanese companies.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/kan-osiris.gif")
```

In the kangaroo plot (Figure `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{fig:kan-osiris-figure}','\\@ref(fig:kan-osiris-gif)'))`) the five grey bars indicate the five categories (Top 25%, 25-50, 50-75, 75-100, Not listed) across years 2006 and 2018. Points represent individual companies with color indicating the country of registration. Points that stay within the bar indicate companies that remain in the same quartile over the entire time period. If a company drops in performance or improves performance to the extent of switching quartile, the corresponding point will jump up or down in this plot. Note that all the companies in the "Not listed" category were listed at some point during this time period. It can be seen that most companies stay in the same quantile group over the time period. There are several companies that move up and down, and most of these are US companies. This supports the @oecd_report claim that US companies have a higher turnover rate compared to Japanese companies.

Perceiving the differences between the movement patterns of groups can be difficult, especially when the groups have different sizes, as is the case here. There are more US companies than Japanese companies, and particularly, there are more US companies in the top 25 percent quantile. Thus, there will likely be more blue points (US) moving than red points (JP) purely for this reason. The difference that is of interest is **relative proportion**, that one country has a higher fraction of companies moving between quartiles than the other. This animated Sankey plot requires the reader to mentally calculate the fraction of points that are moving, separately for the blue and red points, to answer the question of whether one group has a higher movement rate. The task is easier if the two groups have a similar number of points.

@@

  In Figure `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{fig:osiris-figure}','\\@ref(fig:osiris-gif)'))`, the focus is placed on companies that exited the market in 2018 that were in the market at some point since 2006. The grey regions show the proportion of companies in each of the categories in 2006 on the left that exited the market by 2018. (explain the difference in proportion and not listed companies). In 2006, the companies represented by points, colored by country, were in one of the quartiles or "Not listed" category. All the companies in the Top 25% were US companies. Any Japanese companies in the Top 25% in 2006 did not exit the market (Not listed). But actually, there were non-Japanese companies in the Top 25% in 2006. It is also interesting that the larger proportion of the companies in the Top 25% are getting de-listed, and the lower the quartiles, the less likely that the companies will exit the market.

  The wallaby plot can be employed to emphasize the high turnover rate observed in the kangaroo plot, as it specifically focuses on a subset, allowing the users to better perceive the **relative proportion**. However, it is also important to note that this plot is confined to only two-time points (start and end), which may not provide a comprehensive view of the entire story.

@@

```{r}
aeles <- aeles |> 
  filter(party != "other")
```

@@

```{r voter-figure, fig.cap = "Four frames from a wallaby plot visualization show how the top party performs in keeping the old voters of different genders. Most voters remain loyal to the party, but a small fraction of voters with roughly equal male-to-female ratio switch primarily to the other major party. Interestingly, the few individuals who identified as neither male nor female overwhelmingly shifted their party affiliations to the Greens.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/animation-voter.png")
```

```{r voter-gif, out.extra="class = 'gif'", fig.cap = "The wallaby plot visualization shows how the top party performs in keeping the old voters of different genders. Most voters remain loyal to the party, but a small fraction of voters with roughly equal male-to-female ratio switch primarily to the other major party. Interestingly, the few individuals who identified as neither male nor female overwhelmingly shifted their party affiliations to the Greens.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/voter.gif")
```

@@

  According to the survey (@lgbtiq), the Greens Party provided a detailed response to all nine priority areas that focus on changing systems and addressing health and well-being disparities among LGBTIQ+ communities. However, the Labor Party only gives a broad statement of commitment to a range of LGBTIQ+ and human rights issues. This includes working with LGBTIQ+ Australians and advocates to ensure equality before the law and full access to Medicare.

@@

Beginning with inspiration drawn from the New York Times article, this package provides tools to generate animations for temporal flow data to study changes in groups over time. The paper describes how to pre-process the data into a format for making several types of plots and animation. It also explains what needs to be considered when making an animation, such as the persistence of objects, how to achieve this using interpolation, and the visualization design choices automated and those available to users.

  The package provides the tools to create animations for simple temporal patterns with a few steps, but these visualization methods have some limitations. It is primarily for communication rather than exploration because one typically knows what the main patterns are. If there are many categories (quantiles), the display will be too crowded. Because the animation involves perceiving patterns in groups of moving points, it can be challenging with unbalanced sample sizes. This requires the reader to mentally calculate the fraction of points that are in motion.

  There are some potential new directions to develop this work. Some applications might benefit by re-parameterization of the x-axis from years to ages (e.g. company ages, time since the company was founded). This can be done by pre-processing before passing to animbook, but it would be more convenient to automate for the user. One could also imagine re-arranging the mapping of variables to the plot and animations, particularly, split on time and animate on the quantiles to focus on movement through the ranks. The New York Times article had examples of this idea. The static funnel web plot is a good place to experiment with this and to further enhance it visually.
